Drought-sensitive trees survive on deep-water access in a tropical forest community

### 2020-08-10 manuscript.rmd
1. Added objective function for calibration 2. Edited discussion text

\
We calibrated the model over 2012-2018 against three key fluxes in Eq. \@ref(eq:water-balance), namely, (i) evapotranspiration (ET, $E_{v} + E_{g}$) from the eddy flux tower (monthly, 2012-2017), (ii) stream discharge for the Conrad Catchment ($q_{over} + q_{drai}$; daily, 2012-2018) and (iii) volumetric water content (VWC; $W_{liq, z}$; Fig. \@ref(fig:hydro-calib); daily, 2015-2018). We collected VWC data (2016-2018) from a point location at the eddy flux tower with Time Domain Reflectometry (TDR) using three horizontal probes at depths 10, 40 and 100 cm and one vertical probe (0-15 cm). TDR data for the vertical probe was calibrated based on measurements of gravimetric soil moisture on several occasions (Fig. \@ref(fig:TDR-calib)) and mean bulk dry density of 0.8 gcm^-3^ (*n* = 5, cylinder volume = 1250 cm^3^). The same calibration equation was used for other probes. We also leveraged existing opportunistic dry season VWC measurements for the entire 50-ha plot (15, 40 and 100 cm in 2015 and 15 & 40 cm in 2016; @Kupers:2019). Among the 5000 parameter ensembles we chose 100 ensembles that best-fit all three fluxes simultaneously (add details...that is those that maximised the sum of regression coefficients between observed and simulated variables. xx add equation) See Table \@ref(tab:param) for best-fit values and ranges for different parameters. We ran ELM-FATES with the chosen 100 ensembles from 1985-2018, with first 5 years used for model spin-up.

@@

### Effective rooting depths at the tree community level
\
By estimating ERD for large trees of `r length(erd.sp)` canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species species-rich tropical forests. Our chosen ERD model is an advancement over the only other inverse model of ERD (or water uptake depth index, @Chitra-Tarak.2018, *n* = 12 species) in terms of physiologically meaningful representation, fewer parameters, corroboration with tree and soil data on $\delta^2H$. The model in @Chitra-Tarak.2018 estimated six species-specific phenomenological parameters: three as root fraction for corresponding lumped vadose zone layers, and three as shape parameters for water uptake vs availability curves. For each species, they calculated daily water uptake successively from three layers--the supply, to satisfy the demand of the daily maximum evapo-transpiration, a function of PET and LAI. They regressed total supply to total demand ratio over census intervals (termed as growth coefficient) against growth (5 data points), and chose values for six parameters that best-fit (highest R^2^, that is, 2 additional parameters). Our inverse ERD model fits only two parameters by way of regressing depth-specific model against 5 growth data points. Our model was constrained on VPD and species leaf vulnerability (PLC) curves, variables that are shown to be physiologically meaningful [add references]. We benefitted from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (LMA~lamina~ and WSG~100^$\circ$^C~) and parameters of PLC curves for which we collected data on a few species (*n* = 14), a 'hard' trait. This allowed us to scale-up PLC curves at the community scale (*n* = 284 species). We find that per cent loss of conductivity was more rapid for soft wooded and low LMA species as $\Psi_{leaf}$ declined (Fig. S\@ref(fig:comm-plc-curves)A & S\@ref(fig:comm-plc-curves)B). Whether such scaling relationships between hard and soft traits hold true for other forests, and their wider applicability, remains to be seen as data on PLC curves are limited in tropical forests, but counter examples exist [see xx, yy, zz; Lawren, could you suggest references? Is our $K_{leaf}$ vs. $\Psi_{leaf}$ data the first reported for a tropical forest?]. We also note the availability of data on $\delta^2H$ on trees and water sources belowground--soil/substrate matrix water in the vadoze zone or free water below the water table--as important for model calibration/validation. The model in @Chitra-Tarak.2018 were not corroborated with such data as existing methods were inadequate to extract water bound to the soil and saprolite found in their dry forest study site (Jean Riotte, pers. comm.). The model of belowground water-availability in this study (ELM-FATES) is far more advanced compared to that used by @Chitra-Tarak.2018 and was calibrated with data on soil moisture dynamics by depth (15, 40, 100 cm), as well as, evapotranspiration and stream discharge. @Chitra-Tarak.2018 was calibrated only on stream discharge and water table levels. We lacked data on water table levels which would have been useful to constrain water availability dynamics in the unobserved soil layers.

### Future Directions for Model Development
\
The trade-offs that we identify between above- and below-ground traits, water environment vertically belowground and mortality are important for representing diversity in Dynamic Global Vegetation Models especially in species-rich forests. The hydraulic efficiency-safety trade-off between high values of maximum leaf or stem conductivity (high $K_{max}$) and cavitation-resistance (low $\Psi_{50}$ or $\Psi_{88}$) is hypothesised although community-wide examples are rare [@Hacke:2006; @Gleason:2016]. We report such a trade-off at both leaf (*n* = 21 species) and branch level (*n* = 27 species). We also found that LMA~lamina~ and WSG~100^$\circ$^C~ were strong predictors of PLC curves for leaf, but not branch (add SI figures). High $K_{max}$ is elsewhere associated with photosynthetic rates with faster growth potential, but we did not find a relationship between $K_{max}$ and growth rates (*n* = 21 species; xx add figure). These relationships between ERD and aboveground hydraulic traits will provide an important insight on how to model the rooting depth and its coordination or trade-off with other traits to better represent the functional diversity in tropical forests.

>>>>

\
We calibrated the model over 2012-2018 against three key fluxes in Eq. \@ref(eq:water-balance), namely, (i) evapotranspiration ($ET \simeq E_{v} + E_{g}$) from the eddy flux tower (2012-2017), (ii) stream discharge for the Conrad Catchment ($q_{run} \simeq q_{over} + q_{drai}$; 2012-2018) and (iii) volumetric water content (VWC, $\theta_{z} \simeq W_{liq, z}$; 2012-2018). We collected VWC data (2012-2018) from a point location at the eddy flux tower with a vertical Time Domain Reflectometry (TDR) probe ($\theta_{0-0.15m}$). This data was calibrated based on measurements of gravimetric soil moisture at the vertical probe (0-15 cm) on several occasions (Fig. \@ref(fig:TDR-calib)) and mean bulk dry density of 0.8 gcm^-3^ (*n* = 5, cylinder volume = 1250 cm^3^). We also leveraged existing opportunistic dry season VWC measurements by depth for the entire 50-ha plot ($\theta_{0.15m}$, $\theta_{0.4m}$ and $\theta_{1m}$ in 2015 and $\theta_{0.15m}$, $\theta_{0.4m}$ in 2016; @Kupers:2019). Beginning in 2016 we additionally installed three horizontal probes at depths of 0.15, 0.4 and 1m, but depth specific calibration equation was not available, so we applied the same calibration equation to these data as that of the vertical probe. However, for VWC calibration we only used data from the vertical probe to capture the long-term dynamics and plot-wide measurements to capture the dynamics by depth. To find parameter ensembles that best-fit all three fluxes simultaneously, we calculated an objective function for each of the 5000 ensembles based on root mean square error (RMSE) between observations and simulations. Let $f(obs,~sim)$ be normalised RMSE (0-1) between observations and simulations, based on monthly sums for $ET$ and $q_{run}$ and daily avarages for $\theta$s, then the objective function ($Z$) is described as:

\begin{equation}
Z = \frac{1}{6}\sum \left(f(ET,~E_{v} + E_{g}) + f(q_{run},~q_{over} + q_{drai}) + f(\theta_{0-0.15m}, {W_{liq,~\bar{z}}})  + f(\theta_{0.15m}, W_{liq,~`r soil.depths[4]`m}) + f(\theta_{0.4m}, W_{liq,~`r soil.depths[6]`m}) + f(\theta_{1m}, W_{liq,~`r soil.depths[8]`m})\right)
(\#eq:obj-func)
\end{equation}

where $W_{liq,~\bar{z}}$ is the daily average of $W_{liq,~z}$ for $z$ = `r paste(soil.depths[1:4], collapse=", ")` m. To ensure calibration with $\theta_{0.1m}$, we first selected ensembles with  $f(\theta_{1m}, W_{liq,~`r soil.depths[8]`m})$ < 0.2 (~300 ensembles) and among them selected 100 ensembles with lowest values for the objective function $Z$. Model simulations corresponding to these 100 best-fit ensembles are shown against observed fluxes in Fig. \@ref(fig:hydro-calib), including VWC from horizontal probes that were not used for calibration. See Table \@ref(tab:param) for the range of best-fit values for different parameters. We ran ELM-FATES with the chosen 100 ensembles from 1985-2018, with first 5 years used for model spin-up.

@@

### Effective rooting depths at the tree community level
\
By estimating ERD for large trees of `r length(erd.sp)` canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species species-rich tropical forests. Our chosen ERD model is an advancement over the only other inverse model of ERD (or water uptake depth index, @Chitra-Tarak.2018, *n* = 12 species) in terms of physiologically meaningful representation, fewer parameters, corroboration with tree and soil data on $\delta^2H$. The model in @Chitra-Tarak.2018 estimated six species-specific phenomenological parameters: three as root fraction for corresponding lumped vadose zone layers, and three as shape parameters for water uptake vs availability curves. For each species, they calculated daily water uptake successively from three layers--the supply, to satisfy the demand of the daily maximum evapo-transpiration, a function of PET and LAI. They regressed total supply to total demand ratio over census intervals (termed as growth coefficient) against growth (5 data points), and chose values for six parameters that best-fit (highest R^2^, that is, 2 additional parameters). Our inverse ERD model fits only two parameters by way of regressing depth-specific model against 5 growth data points. Our model was constrained on VPD and species leaf vulnerability (PLC) curves, variables that are shown to be physiologically meaningful [add references]. We benefitted from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (LMA~lamina~ and WSG~100^$\circ$^C~) and parameters of PLC curves for which we collected data on a few species (*n* = 14), a 'hard' trait. This allowed us to scale-up PLC curves at the community scale (*n* = 284 species). We find that per cent loss of conductivity was more rapid for soft wooded and low LMA species as $\Psi_{leaf}$ declined (Fig. S\@ref(fig:comm-plc-curves)A & S\@ref(fig:comm-plc-curves)B). Whether such scaling relationships between hard and soft traits hold true for other forests, and their wider applicability, remains to be seen as data on PLC curves are limited in tropical forests, but counter examples exist [Lawren, could you suggest references? Also, folks, is our $K_{leaf}$ vs. $\Psi_{leaf}$ data the first reported for a tropical forest? or could you add how rare it is?]. We also note the availability of data on $\delta^2H$ on trees and water sources belowground--soil/substrate matrix water in the vadoze zone or free water below the water table--as important for model calibration/validation. The model in @Chitra-Tarak.2018 were not corroborated with such data as existing methods were inadequate to extract water bound to the soil and saprolite found in their dry forest study site (Jean Riotte, pers. comm.). The model of belowground water-availability in this study (ELM-FATES) is far more advanced compared to that used by @Chitra-Tarak.2018 and was calibrated with data on soil moisture dynamics by depth (15, 40, 100 cm), as well as, evapotranspiration and stream discharge. @Chitra-Tarak.2018 was calibrated only on stream discharge and water table levels. We lacked data on water table levels which would have been useful to constrain water availability dynamics in the unobserved soil layers.

### Future Directions for Model Development
\
The trade-offs that we identify between above- and below-ground traits, water environment vertically belowground and mortality are important for representing diversity in Dynamic Global Vegetation Models especially in species-rich forests. The hydraulic efficiency-safety trade-off between high values of maximum leaf or stem conductivity (high $K_{max}$) and cavitation-resistance (low $\Psi_{50}$ or $\Psi_{88}$) is hypothesised although community-wide examples are rare [@Hacke:2006; @Gleason:2016]. We report such a trade-off at both leaf (*n* = 21 species) and branch level (*n* = 27 species). We also found that LMA~lamina~ and WSG~100^$\circ$^C~ were strong predictors of PLC curves for leaves, but not for xylem (not shown). Acquisitive traits such as high $K_{max}$ are expected to be associated with faster growth rates. We found a positive but not  significant relationship of adult growth rates ($\ge$ 10cm dbh) with $K_{max, ~leaf}$ (*R*^2^ = 0.13; *p* = 0.126, *n* = `leaf.traits.sp`) and no relationship with $K_{max, ~stem}$ (*n* = `stem.traits.sp` species). The relationships between ERD and aboveground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits in order to better represent the functional diversity of tropical forests.
